---
title: "Hands-on Exercise 2: Thematic Mapping and GeoVisualisation with R"
date: "`r Sys.Date()`"
author:
  - name: Novia Antony
description: |
  welcome to my second hands-on exercise where....

format:
  html:
    toc: true
---
## 1.0 Overview

## 2.0 Setup

### 2.1 Setting Up Your File Directory

..explain later

### 2.2 Data Sources

-   Master Plan 2014 Subzone Boundary (Web) from data.gov.sg
-   Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2011-2020 from Department of Statistics, Singapore

### 2.3 Installation of Packages

In this exercise, we're focusing on the following R packages:

- tmap: This is our go-to package for crafting thematic maps, including types like choropleths and bubble maps.
For a comprehensive analysis, we'll also revisit packages from our previous lesson:

- sf: A key package for loading, handling, and manipulating geospatial data.
- tidyverse: A collection of packages essential for data import, manipulation, and visualization, among other data science activities.

Here's a quick script to ensure we have all these packages ready:

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## 3.0 Importing Data into R

### 3.1 Importing Geospatial Data

```{r}
mpsz <- st_read(dsn = "data/geospatial", 
                layer = "MP14_SUBZONE_WEB_PL")
mpsz
```

### 3.2 Importing Attribute Data

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
```

## 4.0 Data Preparation

### 4.1 Data Wrangling

## 5.0 Choropleth Maps

## 6.0 Final Reflections





