---
title: "Hands-on Exercise 2: Thematic Mapping and GeoVisualisation with R"
date: "`r Sys.Date()`"
author:
  - name: Novia Antony
description: |
  welcome to my second hands-on exercise where....

format:
  html:
    toc: true
---
## 1.0 Overview
..explain later

## 2.0 Setup
..explain later

### 2.1 Setting Up Your File Directory

..explain later

### 2.2 Data Sources

-   Master Plan 2014 Subzone Boundary (Web) from data.gov.sg
-   Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2011-2020 from Department of Statistics, Singapore

### 2.3 Installation of Packages

In this exercise, we're focusing on the following R packages:

- tmap: This is our go-to package for crafting thematic maps, including types like choropleths and bubble maps.
For a comprehensive analysis, we'll also revisit packages from our previous lesson:

- sf: A key package for loading, handling, and manipulating geospatial data.
- tidyverse: A collection of packages essential for data import, manipulation, and visualization, among other data science activities.

Here's a quick script to ensure we have all these packages ready:

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## 3.0 Importing Data into R

### 3.1 Importing Geospatial Data

```{r}
mpsz <- st_read(dsn = "data/geospatial", 
                layer = "MP14_SUBZONE_WEB_PL")
mpsz
```

### 3.2 Importing Attribute Data

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
```

## 4.0 Data Preparation

### 4.1 Data Wrangling

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup()%>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

### 4.2 Joining the attribute data and geospatial data

```{r}
popdata2020 <- popdata2020 %>%
  mutate_at(.vars = vars(PA, SZ), 
          .funs = list(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)

mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))

write_rds(mpsz_pop2020, "chap02/data/rds/mpszpop2020.rds")

```

## 5.0 Choropleth Mapping Geospatial Data with tmap

### 5.1 Thematic Plot with qtm

`tmap_mode("plot")` creates a static map while `tmap_mode("view")` creates an interactive map

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, 
    fill = "DEPENDENCY")
```

### 5.2 Creating a choropleth map by using tmap’s elements

```{r}

tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "Dependency ratio") +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))

```

#### 5.2.1 Drawing a base map

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

#### 5.2.2 Drawing a choropleth map using tm_polygons()

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY")
```


tm_fill() shades the polygons by using the default colour scheme and tm_borders() adds the borders of the shapefile onto the choropleth map.


```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```


to add the boundary of the planning subzones, tm_borders will be used as shown in the code chunk below.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY") +
  tm_borders(lwd = 0.1,  alpha = 1)
```


### 5.3 Data classification methods of tmap


### 5.4 Colour Scheme

### 5.5 Map Layout

### 5.6 Mappping Spatial Object Meeting a Selection Criterion

## 6.0 Final Reflections




